---
title: "HW8"
output: md_document
date: "2026-04-30"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load Packages

```{r}
library(readxl)
library(AnnotationDbi)
library(org.Hs.eg.db)
library(gprofiler2)
library(tidyverse)
```

# Attempted Reproduction

We will be attempting to reproduce the gene-set analysis using [MAGMA](https://cncr.nl/research/magma/) from the "Gene-based analyses suggest an enrichment of IEI genes" portion of the paper. The following steps with code are provided and exact links are contained with the scripts of this repository and may be fetched with the `get_data.sh` script

In regards to simplifications and assumptions, this came down to gene IDs. We could not guarantee the same version of the IEI gene list or the GRCh37 genes. We could have attempted to merge these on a common ID reference like ENTREZ, but this seemed like a time-consuming complexity beyond the need of this assignment. Overall, though, we were generally able to reproduce this MAGMA analysis pipeline. We will discuss discrepencies between our results and theirs in a following section. We ran two tests with MAGAM, one for the IEI gene set list and one for 14 genes that cause monogenic EBV-driven lymphoproliferative diseases.

We also continued on from these results to generate a figure of GO Biological Process annotations. While the paper did not specify how they did these, we will proceed with the GO annotations using `gprofiler2` and `ggplot`.

### 1. Load the data and MAGMA binary

We need to load:
- The magma standalone binary (since this isn't loaded on the cluster)
- The GWAS summary data from Locus Zoom
- The European LD data
- The Gene location file for GRCh37
- The IEI gene set

NOTE: there is a chance that the gene IDs from the IEI gene set and GRCh37 have diverged over time or require mapping. For the sake of this project, we have decided to assume they match closely and will not try to manually ensure all of these map appropriately. We are just going to mark this up to issue with reproducibility and how versioning is a nightmare in gene naming/identification.

```{bash, eval=FALSE}
if [ -d data ]; then
    echo "Directory exists, deleting now..."
    rm -rf data
fi
if [ -d magma ]; then
    echo "Directory exists, deleting now..."
    rm -rf magma
fi
./get_data.sh
```

### 2. Cleanup the GWAS Summary Data from LocusZoom

We need to split the LocusZoom summary data into two files for input into MAGMA:
- a SNP location file for the annotation step
- a SNP P-value file for the Gene-level analysis
- also are removing the HMC region as they do in the paper

```{r, eval=FALSE}
# load the summary data from LocusZoom
lz_stats = read_table("data/summary_stats")

# clean the names
lz_stats = lz_stats %>% 
  rename(
    SNP = rsid,
    CHR = `#chrom`,
    BP = pos
  ) %>% 
  mutate( # we have -log10 p-value, need unadjusted
    P = 10^(-neg_log_pvalue)
  )

# filter our the Major Histocompatibility Complex (HMC) region
# chr6:28,510,120–33,480,577
# https://www.sciencedirect.com/science/article/pii/S1357272520301990
lz_stats = lz_stats %>% 
  filter(
    !(CHR == 6 & BP >= 28510120 & BP <= 33480577)
  )

# create the file for annotation
write.table(
  lz_stats %>% select(SNP, CHR, BP),
  file = "data/snp.loc",
  sep = "\t",
  row.names = FALSE,
  quote = FALSE
)

# create the file for analysis
write.table(
  lz_stats %>% select(SNP, P),
  file = "data/gwas.txt",
  sep = "\t",
  row.names = FALSE,
  quote = FALSE
)

```

### 3. Prepare the IEI Gene Set File
```{r}
gene_table = read_xlsx("data/IEI.xlsx")

# ignoring any genes with duplicate IDs or other text

gene_table$Gene = gene_table$`Genetic defect`

iei_genes = gene_table %>% 
  filter(
    !str_detect(Gene, "\\s")
  )

iei_genes = unique(trimws(iei_genes$Gene))

# convert these to entrez IDs
iei_genes <- mapIds(
  org.Hs.eg.db, 
  keys = iei_genes, 
  column = "ENTREZID", 
  keytype = "SYMBOL", 
  multiVals = "first"
)

# remove any NAs (failed to map to ENTREZ ID)
iei_genes = iei_genes[!is.na(iei_genes)]

iei_genes = c(
  "IEI",
  iei_genes
)

# the set of "14 genes that cause monogenic EBV-driven lymphoproliferative diseases"
# cited from this paper: https://link.springer.com/article/10.1007/s00439-020-02145-3
# the paper was not too clear, so I had Claude generate this list and checked it
lymph_genes = c(
  "SH2D1A",
  "XIAP",
  "ITK",
  "MAGT1",
  "CD27",
  "CD70",
  "CTPS1",
  "CORO1A",
  "RASGRP1",
  "TNFRSF9",
  "STK4",
  "CARMIL2",
  "DOCK8",
  "PRKCD"
)

# convert these to entrez IDs
lymph_genes <- mapIds(
  org.Hs.eg.db, 
  keys = lymph_genes, 
  column = "ENTREZID", 
  keytype = "SYMBOL", 
  multiVals = "first"
)

# remove any NAs (failed to map to ENTREZ ID)
lymph_genes = lymph_genes[!is.na(lymph_genes)]

lymph_genes = c(
  "Lymph_Disease",
  lymph_genes
)

# print these out to a file
cat(
  iei_genes,
  file = "data/IEI_geneset.txt",
  sep = "\t"
)
cat(
  "\n",
  file = "data/IEI_geneset.txt",
  sep = "",
  append = TRUE
)
cat(
  lymph_genes,
  file = "data/IEI_geneset.txt",
  sep = "\t",
  append = TRUE
)

```

### 4. Run the MAGMA Pipeline

1. Run SNP to Gene Annotation with MAGMA
2. Run Gene-level Analysis with MAGMA
3. Run the Gene Set Enrichment with MAGMA

```{bash, eval=FALSE}
./run_magma.sh
```

### 5. Overview of the Results with a Figure

Recreation of "Fig. 3: Characterization of non-MHC risk loci associated with EBVread+." Section "b":

```{r}
# load the genes from IEI
magma_genes = read.table(
  "data/ebv_gwas.genes.out",
  header=TRUE
)

# Number of genes tested
n_genes <- nrow(magma_genes)
n_genes

# Bonferroni threshold
bonferroni_threshold = 0.05 / n_genes
bonferroni_threshold

# Add corrected p-value column
magma_genes$p_bonferroni = p.adjust(magma_genes$P, method = "bonferroni")

# Significant genes
sig_genes = magma_genes[magma_genes$p_bonferroni < 0.05, ]
sig_genes = sig_genes$GENE

results <- gost(
  query = sig_genes,
  organism = "hsapiens",
  sources = "GO:BP",
  correction_method = "bonferroni",
  evcodes = TRUE,
  numeric_ns = "ENTREZGENE_ACC"   # tells g:Profiler these are Entrez IDs
)

# get the top 30, same number displayed as the paper
res_df = results$result[order(results$result$p_value),
                          c("term_name", "p_value", "term_size",
                            "intersection_size", "intersection")]

top_30 = head(res_df, 30) %>% 
  mutate(
    `-log10(P)` = -log10(p_value)
  ) %>% 
  arrange(
    `-log10(P)`
  )
nrow(top_30)

# generate the plot from the paper
top_30 %>% 
  ggplot() +
  aes(
    x = `-log10(P)`,
    y = reorder(term_name, `-log10(P)`)
  ) +
  geom_col() +
  geom_vline(
    xintercept = -log10(bonferroni_threshold),
    color = "red",
    linetype = "dashed",
    size = 1
  ) +
  theme_minimal() +
  theme(axis.title.y = element_blank())

# check which are below our bonferroni threshold
top_30_sig = top_30 %>% 
  filter(
    `-log10(P)` > -log10(bonferroni_threshold)
  )
nrow(top_30_sig)
top_30_sig

```

# Our Reproduction vs Original Paper

Our results are found in the `./data/ebv_gwas_IEI.gsa.out`.

## Their Results

### IEI Genes

Given the p-value is less than 0.05, they can say that the IEI genes are enriched compared to non-IEI genes by 0.19 units.

- P = 4.66 × 10−6
- beta = 0.19
- s.e.m. (AKA SE) = 0.04
- Number of Genes = 456
- Bonferroni Corrected P < 6.5 × 10−6; n = 7,743

### Genes that cause monogenic EBV-driven lymphoproliferative diseases

Given the p-value is greater than 0.05, they conclude that say that the monogenic EBV-driven lymphoproliferative disease related genes are not enriched compared to other genes. They do note, however, that the effect size for these genes are higher than the 

- P = 0.055
- beta = 0.35
- s.e.m. (AKA SE) = 0.22
- Number of Genes = 14

## Our Results:

### IEI Genes

We come to the same conclusion as the paper with a higher p-value and lower beta, signifying overall less impressive results for the IEI genes. This may be due to the 23 missing genes from our set.

VARIABLE      TYPE  NGENES         BETA     BETA_STD           SE            P
IEI            SET     433      0.14181     0.021631     0.044734   0.00076351

Bonferroni Corrected P < 2.75e-06; n = 18,167

### Genes that cause monogenic EBV-driven lymphoproliferative diseases

This is by far the biggest difference. Our results for the monogenic EBV-driven lymphoproliferative disease related genes not only reached statistical significance, but also lead to a much higher beta of about 0.61.

VARIABLE           TYPE  NGENES         BETA     BETA_STD           SE            P
Lymph_Disease       SET      11      0.60817     0.014961      0.28641     0.016866

## Conclusion

While the original paper landed on 30 GO:BP results and appears to have chosen to show the top 30, our analysis only returned 22 biological process, of which, only 5 were significant:

- chromatin organization
- protein-DNA complex assembly
- protein-DNA complex organization
- nucleosome assembly
- nucleosome organization

While the original paper appeared to return results that lead towards the importance of many different types of immune cells, our results are more around how EBV alters the host epigenetically. These results are consistent with a the [literature](https://pubmed.ncbi.nlm.nih.gov/40598900/) and also [here](https://www.youtube.com/watch?v=dQw4w9WgXcQ).,

# AI Usage

My AI usage for this assignment can be summed up as moderately helpful. I mostly began with Claude, then ChatGPT, followed by Google's AI feature in the search engine (as I ran out of free prompts). Overall, the LLMs failed at providing links to the actual data for download. The `get_data.sh` script is meant to initialize all required data and the magam binary. While Claude was able to guess for a pretty close script, I still had to manually confirm each data file and how to successfully decompress them. After getting a general sense of the pipeline to be run from Claude, I moved over to ChatGPT for actual skeleton scripts for running each of the three magma commands of this pipeline and to confirm the actual schema of the data (input/output and specific column names of these) of each step. Generally, Google was mainly used just to answer small coding questions that I could not remember off the top of my head. In comparing our results to the original paper, AI was also useful in trying to find the columns and files out our outputs that matched the paper. I eventually went back to Claude to help confirm the 14 genes related to monogenic EBV-driven lymphoproliferative diseases as neither this paper nor the cited paper gave a clear and concise list at to which genes these were. Lastly, I asked Claude how to generate the figure of GO terms and was directed to use the `gprofiler2` R package. The paper did not distinguish how they mapped to the GO biological process ontology.

# Division of Work

Shane did the "Vibe-Replication (Code + Explanation)" and Kash did the "Paper Summary (brief)" and "Rationale, Methods, and Results". Worked in a "divide and conquer" followed by a get together where we taught each other our respective contributions. No disagreements to mention.

## Understanding and Difficulties

### Shane

Overall, I had issues trying to get the data in a clear and consistent way. Everyone seems to have their own way of storing data as some random link on some website somewhere. I understand now the pain in reproducing these studies. There really should be some kind of constant way of storing data/code for reproduction. I know that there is a tool called [Datalad](https://www.datalad.org/) that is pushing for this. I feel like what I understood was using AI to code and rapidly (or as quicker than before AI) to ge to a solution. Honestly, AI does kind of make these replication do-able in a day or two whereas in the past it would be a much larger project.

### Kash

